---
title: "MP2"
author: "Michael Bevilacqua, Colin Flaherty, Phil Kaplan"
date: "2024-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(knitr)
```

#Introduction 
The purpose of this mini project was to examine the force output and maximum load that can be supported by the forelimb. Skeletal geometry and measured parameters including muscle length and mass were used to create a model to predict the force and torque produced by the human bicep over a range of provided extension/flexion angles. The predicted maximum mass that can be curled according to the model was tested in the gym to assess validity of the model. Insert some studies with relevant findings. 


#Methods

Bicep muscle mass (Mmuscle) was estimated by shaping plato to the size of the bicep and weighing the plato model. Plato's density is similar to that of muscle. Arm length (Larm), Inlever length (Li), and Outlever length (Li) were obtained for each participant using a meter stick. Using the skeletal geometry of the human forelimb and the previously recorded measurements,  Muscle lengths (Lmuscle) at each flexion angle and Muscle insertion angle at each flexion angle(θ) were calculated. Finally, an R function was created using previously measured or calculated parameters to determine muscle torque (muscle_torque) and the maximum amount of mass that could be theoretically curled at each angle (max_mass).


```

```{r data, message=FALSE, echo=FALSE, warning=FALSE}
dat1 <- tibble(
 Person= c("Phil", "Colin", "Michael"),
  Mmuscle = c(0.418, 0.252, 0.344),
  Lo = c(0.320, 0.280, 0.320),
  Li = c(0.080, 0.065, 0.080),
  Larm = c(0.350, 0.31, 0.305)
)
print(dat1)
```

```{r, message=FALSE, echo=FALSE, warning=FALSE}
#Writing Needed Functions 

#Insertion angles
law_cos <- function(o, l, r) {
  acos((-o^2 + l^2 + r^2) / (2 * l * r))
}


law_cos2 <- function(l, r, theta) {
  sqrt(l^2 + r^2 - 2 * l * r * cos(theta))
}

#Establishing rho, k, and g 
rho <- 0.00105
k <- 80
g <- 9.81 

ffl <- function(Lmuscle, Lmuscler) {
  -6.25 * ((Lmuscle / Lmuscler)^2) + 12.5 * (Lmuscle / Lmuscler) - 5.25
}

# Function to calculate PCSA and Fmax
PCSA <- function(Mmuscle, phi, Lmuscler) {
  Mmuscle * cos(phi) / (rho * Lmuscler)
}
PCSA

```

```{r functions, message=FALSE, echo=FALSE, warning=FALSE}
model <- function(mass, Lmuscler, Lmuscle, theta, Lo, rho = 0.00105, k = 80) {
  Lmuscler_cm <- Lmuscler * 100 
  Lmuscle_cm <- Lmuscle * 100  
  Lo_cm <- Lo * 100  
  
  pcsa <- PCSA(mass, 0, Lmuscler_cm) 
  Fmax <- pcsa * k  
  
  ffl_value <- ffl(Lmuscle_cm, Lmuscler_cm * 0.8)  
  
  Fmuscle <- Fmax * ffl_value 
  
  torque <- Fmuscle * sin(theta) * Lo_cm 
  
  return(torque)
}
```

```{r, message=FALSE, echo=FALSE, warning=FALSE, results = 'hide' }

flexion_angles <- c(50, 75, 100, 125, 150) * pi / 180

dat_updated <- dat1 %>%
  crossing(flexion = flexion_angles) %>%
  mutate(
    Lmuscle = law_cos2(Larm, Li, flexion),
    theta = law_cos(Larm, Li, Lmuscle)
  ) %>%
  mutate(
    Lmuscler = ifelse(flexion == (150 * pi / 180), Lmuscle, NA_real_)
  )

dat_with_torque <- dat_updated %>%
  rowwise() %>%
  mutate(
    muscle_torque = model(Mmuscle, Lmuscler, Lmuscle, theta, Lo),
    max_mass = muscle_torque / (Lo * 100 * g)
  ) %>%
  ungroup()

print(dat_with_torque)

#Finding Lmuscler
reference_Lmuscler <- dat_with_torque %>%
  filter(flexion == 150 * pi / 180) %>%
  select(Person, Lmuscler) %>%
  distinct()

#Changing Lmuscler values across all rows belonging to each person
dat_with_torque <- dat_with_torque %>%
  left_join(reference_Lmuscler, by = "Person", suffix = c("", "_ref")) %>%
  mutate(
    Lmuscler = ifelse(is.na(Lmuscler), Lmuscler_ref, Lmuscler)
  ) %>%
  select(-Lmuscler_ref) %>%
  mutate(
    muscle_torque = model(Mmuscle, Lmuscler, Lmuscle, theta, Lo),
    max_mass = muscle_torque / (Lo * 100 * g)
  )


print(dat_with_torque)
```

In the following table predicted Tmuscle and Mload values for each flexion angle are provided. 
```{r}
kable(dat_with_torque, caption = "predicted τmuscle and Mload values at each flexion angle")
```



```{r}
ggplot(dat_with_torque, aes(x = flexion * 180 / pi, y = muscle_torque, color = Person)) +
  geom_line() +
  labs(x = "Flexion Angle (degrees)", y = "Muscle Torque (Nm)", title = "Muscle Torque Across Flexion Angles") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")

ggplot(dat_with_torque, aes(x = flexion * 180 / pi, y = max_mass, color = Person)) +
  geom_line() +
  labs(x = "Flexion Angle (degrees)", y = "Maximum Load (kg)", title = "Maximum Load Across Flexion Angles") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```

#Results
The largest maximum load was obtained for each participant at an angle of 75 degrees. These findings suggest that at a 75 degree angle the bicep is near its optimal length where the greatest number of cross-bridges can be formed. As seen in the graph, as the flexion angle increases (>75 degrees) the maximum load decreases. This is consistent with expectations given that as the angle of the joint increases, the bicep muscle length increases and the amount of torque it can produce decreases. 


